Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-05 paper.tex
non it theorems etc

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

@@

\begin{definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a dependent
  pair of shapes, and positions indexed by those shapes. 
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{definition}
We often use containers to define types like lists or trees in a more generic
way.
For instance, while lists would usually be defined like so:
\begin{equation}
  \begin{aligned}
    \mathbf{List}(A) &= \\
      & \left| \; \text{nil} : \mathbf{List}(A)   \right; \\
      & \left| \; \text{cons} : A \rightarrow \mathbf{List}(A) \rightarrow \mathbf{List}(A) \right;
  \end{aligned}
\end{equation}
We could equivalently define them as a container:
\begin{equation}
  \mathbf{List}(A) = \left\llbracket \mathbb{N} , \mathbf{Fin} \right\rrbracket
\end{equation}

The two representations are equivalent, and they lead to two versions of split
enumerability which are equivalent.
\begin{lemma} \label{container-split-enum-equiv}
  The container-based definition of split enumerability is equivalent to the
  inductive type version.
  \begin{equation}
    \sum_{(\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket A)} , \prod_{(x : A)} , x \in xs \simeq
    \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
\end{lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

>>>>

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsection{Containers}
One last part is missing from the finiteness definition above: the definition of
\(\mathbf{List}\).
The standard definition is as follows:
\begin{definition}[Lists as an Inductive Type]
  \begin{equation}
    \begin{aligned}
      \mathbf{List}(A) &= \\
      & \left| \; \text{nil} : \mathbf{List}(A)   \right; \\
      & \left| \; \text{cons} : A \rightarrow \mathbf{List}(A) \rightarrow \mathbf{List}(A) \right;
    \end{aligned}
  \end{equation}
\end{definition}

However, as we will see, it will be occasionally more convenient to work with an
equivalent type defined as a \emph{container}.
\begin{definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a dependent
  pair of shapes, and positions indexed by those shapes. 
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{definition}

We often use containers to define types like lists or trees in a more generic
way.
While they tend to be more cumbersome to work with (especially for termination
checking) they have a number of distinct advantages.
In particular, when finiteness predicates are defined in terms of containers,
certain proofs (we will point them out as we come upon them) become trivial.

Happily, we can have our cake and eat it too.
Since we have access to the univalence axiom, we can transport along any
equivalence to a more convenient representation with good justification.

\begin{lemma}
  The two definitions of lists---the traditional inductive type, and the
  container---are equivalent.
  \begin{equation}
    \mathbf{List} \simeq \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{lemma}
\begin{proof}
\todo[inline]{Agda code for proof?}
\end{proof}

\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

@@
